Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-13 numerical.tex
updating

\section{Numerical solutions}\label{numerical}
In this section we numerically solve the linear stability problem
corresponding to Eq. \ref{lin_mass}---\ref{lin_energy}. 
We relax the thin-disk, low-frequency and nearly-Keplerian
approximations used in the analytical discussion above, and  
use the full expressions of $\rho$, $\Omega$ and
$\kappa$ given in \S\ref{eqm}. We also set $\hat{g}_c=1$ to
account for the background radial disk structure.

@@

Unless otherwise stated, our fiducial disk model is: a nearly 
vertically isothermal disk with $\Gamma=1.011$, stable
stratification $\gamma=1.4$, $(p,q,\epsilon)=(-1.5,-1,0.05)$, and 
vertical domain size $\zmax=5H$. In Fig. \ref{omega_z} we plot the
corresponding equilibrium rotation profile, which shows that $\Omega$
decreases slightly away from the mid-plane.

@@

\subsection{VSI with rapid thermal relaxation}\label{vertiso_pertiso} 
We first calculate the VSI in a disk with rapid thermal relaxation by
setting $\beta=10^{-3}$. (Smaller values yield the same results.) This
is similar to strictly isothermal perturbations in a vertically
isothermal disk considered in \cite{nelson13} and \cite{mcnally14}. % We
% compare results with our analytical discussion of isothermal 
% perturbations in Appendix \ref{iso_discuss}.

Fig. \ref{iso_eigen_kx} compares the numerically obtained eigenvalues 
for the fundamental mode and that calculated from 
Eq. \ref{simple_growth} (with $L=1$). We find good agreement in the
wave frequency $\omega$ at all $\khat$ and in growth rates $\nu$ for
$\khat\gtrsim 10$. Growth rates are over-estimated by our analysis but 
are $O(\epsilon\Omega_k)$, which is consistent with the discussion in
Appendix \ref{max_growth1}. The match in $\nu$ worsens for
decreasing $\khat$ due to the neglect of global radial structure in
the analysis. Nevertheless, given the number of approximations made
to calculate growth rates analytically, this agreement is
satisfactory, especially at large $\khat$.

@@

Fig. \ref{lowfreq_eigenfunc} plots the eigenfunctions $W$ and $\delta 
v_z$ for the fundamental mode with $\khat=10$ (i.e. the mode with
maximum growth rate in Fig. \ref{iso_eigen_kx}). We find $W\propto z$
throughout most of the disk, corresponding to a constant vertical
velocity perturbation, as expected for the fundamental mode according
to explicit solutions developed in Appendix \ref{iso_explicit}. We
show a meridional visualization of this mode in
Fig. \ref{lowfreq_eigenfunc_2d}. The $x$ axis is stretched for
clarity; radial velocities are in fact typically much
smaller than vertical velocities. Most of the meridional kinetic energy is
contained within $\sim 2H$ of the midplane due to the density 
stratification.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/eigenvectorW_iso} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  1cm]{figures/eigenvectorvz_iso}
  \caption{Pressure perturbation $W$ (top) and vertical velocity
    perturbation $\delta v_z$ of the fundamental VSI for a
    perturbation  wavenumber $\khat=10$  The real 
    (imaginary) parts of $W$ and $\delta v_z$ are plotted as solid
    (dotted) lines. 
    \label{lowfreq_eigenfunc}
  }
\end{figure}

\begin{figure}
%  \includegraphics[scale=0.345,clip=true,trim=0cm 0cm 2.5cm
%  0cm]{figures/result2d_vel}\includegraphics[scale=0.345,clip=true,trim=1.9cm 0cm 0cm
%  0cm]{figures/result2d_mom} 
  \includegraphics[width=\linewidth]{figures/result2d_iso}
  \caption{Visualization of the fundamental VSI mode in
    Fig. \ref{lowfreq_eigenfunc}. The color scale corresponds to the
    pressure perturbation $\delta P=\rho W$ scaled by its maximum value.
    Arrows correspond to the vector field field $\sqrt{\rho}(\delta
    v_x,\delta v_z)$. Note that the horizontal axis has been scretched
    for clarity.  
    \label{lowfreq_eigenfunc_2d}
  }
\end{figure}

\subsubsection{Case study: $\khat=10$}
In Fig. \ref{compare_modes_iso_kx10} we plot eigenvalues of unstable
modes with $\khat=10$. The result for our fiducial setup is shown
in black dots. We also plot other calculations with different
vertical boundary conditions: rigid boundary and $\zmax=5H$ (red
cross),  free surface and $\zmax = 3H$ (blue triangles), free surface
and $\zmax=7H$ (green cross).

The fundamental mode, marked by `f', corresponds to the eigenvalue
with the smallest $|\sigma|$. We see that the fundamental mode is
robust against vertical boundary conditions as it appears at the same
position in all cases. Higher order modes with larger growth rates
(i.e. the branch where $\nu$ increases with $|\omega|$) require larger
vertical domains.  The `tail' end with decreasing $\nu$ to the right
of the plot differ significantly between the cases, and are thus
likely associated with boundary conditions.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_modes_iso_kx10.ps}
  \caption{Unstable modes for a disk with 
    $(\gamma,\Gamma)=(1.4,1.011)$, $(p,q,\epsilon)=(-1.5,-1,0.1)$ and
    $\beta=10^{-3}$. The perturbation wavenumber is $\khat=10$. 
    The fiducial setup with a free surface ($\Delta P=0$) and 
    vertical domain size $\zmax=5H$ is shown by black dots Other
    cases correspond to different boundary condtions applied at
    different heights. The fundamental mode is marked by `f'.   
    \label{compare_modes_iso_kx10} 
  }
\end{figure}

@@

To illustrate this, we plot the mode diagram for $\khat=30$ in  
Fig. \ref{compare_modes_iso_kx30} with different boundary 
conditions. Surface modes are readily identified for 
$\zmax\geq 5H$ and are marked by `s'. These modes have
already been discussed in the aforementioned studies. They typically
have the largest growth rates but require sufficiently large vertical
domain size and their eigenvalues vary with $\zmax$.

Due to their dependence on boundary conditions, we 
do not consider surface modes further in detail. In fact, we will see
in the next section that surface modes are more effectively stabilized
by finite thermal relaxation than the fundamental mode. %  Thus, surface
% modes may not be the most relevant in realistic disks.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_modes_iso_kx30_tags.ps}
  \caption{Same as Fig. \ref{compare_modes_iso_kx10} but for $\khat=30$
    \label{compare_modes_iso_kx30}. Examples of surface modes are
    marked with `s'. These are found for sufficiently large vertical
    domain sizes. 
  }
\end{figure}

\subsection{Effect of thermal relaxation}\label{therm_relax_eff}
We now consider the effect of thermal relaxation by increasing
$\beta$. Note that for the standard disk model the critical thermal 
timescale $\beta_\mathrm{crit} = 0.125$.

Fig. \ref{compare_modes_cool_kx10} show unstable modes with $\khat=10$
and $\beta\in[0.01,0.1]$. In addition to our fiducial setup with 
$\zmax=5H$, we also plot results for $\zmax=3H$ and $\zmax=7H$. In all
three cases the number of unstable modes decrease with increasing
$\beta$. The fundamental mode is not the most unstable for 
$\beta=0.01$ as there is a large number of higher order modes, but
these depend on boundary conditions.

However, the fundamental mode is dominant for $\beta \geq 
0.03$. Higher order modes, with larger growth rates than the
fundamental mode, are more effectively stabilized with increasing 
$\beta$.

For $\beta=0.1$, which is close to $\beta_\mathrm{crit}$, we
only find the fundamental mode.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/compare_modes_cool_kx10_z3.ps} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/compare_modes_cool_kx10_z5.ps}
  \includegraphics[width=\linewidth]{figures/compare_modes_cool_kx10_z7.ps}
  \caption{Unstable modes with $\khat=10$ and thermal
    relaxation timescales $\beta=0.01$ (black dots), $\beta=0.03$ (red
    crosses), $\beta=0.05$ (green triangles) and $\beta=0.1$ (blue
    crosses); for different vertical domain sizes $\zmax=3H$ (top),
    $\zmax=5H$ (middle, standard setup) and $\zmax=7H$ (bottom). The
    fundamental mode is marked with `f'. 
    \label{compare_modes_cool_kx10} 
  }
\end{figure}

Fig. \ref{compare_modes_cool_kx30} show unstable modes with
$\khat=30$. In this case surface modes appear for sufficiently large   
$\zmax$. However, they are effectively stabilized with increasing 
$\beta$. For $\zmax=3H$, the fundamental mode becomes dominant for
$\beta\geq 0.03$, as before. For $\zmax=5H$ and $\zmax=7H$, we find
the appearence of modes slightly more unstable than the fundamental
mode even when $\beta=0.1$ (the leftmost eigenvalues in the lower two
panels). However, these modes are clearly associated with the vertical
domain size as they are absent for $\zmax=3H$. % Thus, for $\zmax=5H$ and 
% $\zmax=7H$  the fundamental mode is likely dominant for 
% $\beta\geq0.05$

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/compare_modes_cool_kx30_z3.ps} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/compare_modes_cool_kx30_z5.ps}
  \includegraphics[width=\linewidth]{figures/compare_modes_cool_kx30_z7.ps}
  \caption{Same as Fig. \ref{compare_modes_cool_kx10} but for
    $\khat=30$. Examples of surface modes are marked with `s', and 
    appear for sufficiently large $\zmax$.
    \label{compare_modes_cool_kx30} 
  }
\end{figure}

Fig. \ref{compare_modes_cool_kx10} and \ref{compare_modes_cool_kx30} 
show that the fundamental mode is not sensitive to boundary
conditions. Thus, tracking fundamental mode is a
more robust way of assessing stability as a function of $\beta$,
rather than searching for the most 
unstable mode, because the latter may be associated with boundary
conditions. In fact, the fundamental mode bcomes dominant 
(or close to it) with increasing $\beta$.

Fig. \ref{lowfreq_eigenfunc_cool}---\ref{lowfreq_eigenfunc_2d_cool}
shows the fundamental mode with $\khat=10$ in a disk with 
$\beta=0.1$. These plots can be compared the case with rapid thermal
relaxation (Fig. \ref{lowfreq_eigenfunc}---\ref{lowfreq_eigenfunc_2d}). We see 
that increasing $\beta$ causes the region characterized by $W\sim z$ and  
$\delta v_z\sim$ constant closer to the mid-plane, with oscillatory
behavior emerging near vertical boundaries where buoyancy becomes
important. This leads to the `tilted' pressure perturbations seen in
Fig. \ref{lowfreq_eigenfunc_2d_cool}
(cf. Fig. \ref{lowfreq_eigenfunc_2d}). % However, most of the meridional
% momentum is still contained within $|z|\lesssim 2H$ as for 
% the case with $\beta=10^{-3}$.

@@

\subsection{Critical thermal relaxation
  timescale}\label{bcrit_num_test}
We now test the critical thermal relaxation timescale given by
Eq. \ref{iso_vsi_cond}. We follow the fundamental mode by starting
from eigenvalues for $\beta=10^{-3}$ and slowly increase the thermal
relaxation timescale up to $\beta=10^3$. Fig. \ref{bcrit_compare1}
show the obtained growth rates, together with the critical 
thermal relaxation timescale $\beta_\mathrm{crit}=0.125$ for our
fiducial disk parameters.

For $5\lesssim\khat\lesssim30$, we find $\beta_\mathrm{crit}$ provides
an accurate prediction of the upper limit to the thermal relaxation 
timescale for the fundamental mode. For smaller $\khat$, 
$\beta<\beta_\mathrm{crit}$ the VSI can operate slightly beyond
$\beta_\mathrm{crit}$ (recall that $\beta_\mathrm{crit}$ was derived
by assuming $\khat\gg 1$). For $\khat\geq 50$, we find perturbations
acquire large amplitudes near the vertical boundaries. Complications
% to our simple solution procedure 
may arise from surface modes. Nevertheless, growth rates rapidly
decrease as $\beta$ approach and increase beyond
$\beta_\mathrm{crit}$.

It is interesting to note that there is a thermal relaxation timescale that
maximizes the mode decay rate for $1\lesssim \khat \lesssim 30$. Here,
it is $\beta\simeq1$---$2$ or about $0.1$---$0.3$ orbital
periods. This is in fact consistent with Fig. 12 of \cite{nelson13}
where the perturbed kinetic energy in their numerical simulations was
observed to decay most rapidly for a thermal timescale of $\sim 0.1$
orbits.

%implications for nonlinear sims?

\begin{figure}
   \includegraphics[width=\linewidth]{figures/gcorr_compare2} 
   \caption{Growth rate of the fundamental VSI in the fiducial disk
     model as a function of the thermal relaxation timescale $\beta$
     The vertical line
     is the critical thermal timescale $\beta_\mathrm{crit}$  obtained
     from Eq. \ref{iso_vsi_cond}. 
     \label{bcrit_compare1}}   
 \end{figure}

% may need to explain why choose k=10
% bcrit independent of k for k>>1 
% for some parameter regimes (e.g. large q, large epsilon) 
% behavior at larger kx and finite bcool not understood. 
We further test the dependence of the critical thermal relaxation timescale 
$\beta_\mathrm{crit}$ on disk parameters
(Eq. \ref{iso_vsi_cond}).  We use the previous case study as a
reference point and vary $q\in[-0.2,-1.2]$, 
$\gamma\in[1.2,2.0]$, and $\epsilon\in[0.02,0.1]$
separately.%  Here, we use a wavenumber $\khat=10$, for which we
% find growth rates reach zero for the range of disk parameters
% considered, so that a numerical $\beta_\mathrm{crit}$ can be defined precisely.

The numerically obtained $\beta_\mathrm{crit}$ is shown in
Fig. \ref{bcrit_compare} in comparison with Eq. \ref{iso_vsi_cond}.
Our numerical results generally agree with Eq. \ref{iso_vsi_cond}. The
agreement improves with decreasing  $|q|$,  $\epsilon$ and increasing
$\gamma$, i.e. weaker instability. There is a noticeable difference
as $\gamma\to1$ because the derivation of $\beta_\mathrm{crit}$
assumed $\gamma>1$.

Fig. \ref{bcrit_compare}, together with the previous section
(Fig. \ref{bcrit_compare1}), shows that $\beta_\mathrm{crit}$ as given
by Eq. \ref{iso_vsi_cond} characterizes the thermal timescale below
which perturbations are effectively isothermal 
so that the VSI operates.

Fig. \ref{bcrit_compare1} shows that $\beta_\mathrm{crit}$ provides a  
practical indication for the maximum thermal relaxation timescale to
permit the fundamental VSI. % From our discussion in 
% \S\ref{therm_relax_eff}, which indicate the fundamental mode typically
% dominant 
\cite{nelson13} found a thermal relaxation timescale of $\beta\gtrsim
0.6$ stabilized the VSI in non-linear hydrodynamic simulations. This 
is consistent with our results.

% From the previous section (Fig. \ref{relax_eigenW_num}) we
% expect that for other wavenumbers,  still gives a characteristic
% thermal timescale below which the perturbations are effectively
% isothermal.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.cm 0cm
  0cm]{figures/bcrit_compare_q.ps} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm
  0.8cm]{figures/bcrit_compare_g.ps}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm
  0.8cm]{figures/bcrit_compare_e.ps} 
  \caption{Dependence of the upper limit to the thermal relaxation timescale
    $\beta_\mathrm{crit}$ for the fundamental VSI mode on disk
    parameters. The fiducial parameters are $\Gamma=1.011$,
    $\gamma=1.4$ and $(p,q,\epsilon)=(-1.5,-1,0.05)$. Top: varying
    $q\in[-1.2,-0.2]$; middle: varying $\gamma\in[1.2,2.0]$; bottom:
    varying $\epsilon\in[0.02,0.1]$. The perturbation wavenumber is
    $\khat=10$.  
    \label{bcrit_compare}}  
\end{figure}

@@

For $\gamma>\Gamma$, i.e. stably stratified disks,
Fig. \ref{gcorr_compare_vnoniso} shows that introducing finite thermal
relaxation rapidly stabilizes the fundamental mode. This is
similar to that observed for nearly vertically isothermal disks
(Fig. \ref{bcrit_compare1}). For $\gamma=1.7,\,2.0,\,2.5$, growth
rates reach zero at  $\beta\simeq0.17,\,0.083,\,0.045$, respectively.
Interestingly, these values are equal to
$\epsilon|q|/(\gamma-\Gamma)$. This suggests that the critical thermal
relaxation timescale for vertically non-isothermal disks can also be
estimated by Eq. \ref{iso_vsi_cond} but with $\gamma-1$ replaced by
$\gamma-\Gamma$.

>>>>

\section{Numerical solutions}\label{numerical}
In this section we numerically solve the linear stability problem
corresponding to Eq. \ref{lin_mass}---\ref{lin_energy}. 
We relax the  low-frequency, nearly-Keplerian and thin-disk 
approximations used in the analytical discussion above. We 
use the full expressions of $\rho$, $\Omega$ and 
$\kappa$ and $c_s$ given in \S\ref{eqm}. We also set $\hat{g}_c=1$ to  
account for the background radial disk structure.

@@

Unless otherwise stated, our fiducial disk model is a nearly 
vertically isothermal disk with $\Gamma=1.011$, stable
stratification $\gamma=1.4$, $(p,q,\epsilon)=(-1.5,-1,0.05)$, and 
vertical domain size $\zmax=5H$. In Fig. \ref{omega_z} we plot the
corresponding equilibrium rotation profile, which shows that $\Omega$
decreases slightly away from the mid-plane.

@@

\subsection{VSI with rapid thermal relaxation}\label{vertiso_pertiso} 
We first calculate the VSI in a disk with rapid thermal relaxation by
setting $\beta=10^{-3}$. (Smaller values yield the same results.) This
is similar to strictly isothermal perturbations in a vertically
isothermal disk considered in \cite{nelson13} and \cite{mcnally14}.

Fig. \ref{iso_eigen_kx} compares the numerically obtained eigenvalues 
for the fundamental mode and that calculated from 
Eq. \ref{simple_growth} (with $L=1$). We find good agreement in the
wave frequency $\omega$ at all $\khat$ and in growth rates $\nu$ for
$\khat\gtrsim 10$. Growth rates are over-estimated by our analysis but 
are $O(\epsilon\Omega_k)$, which is consistent with the discussion in
Appendix \ref{max_growth1}. The match in $\nu$ worsens for
decreasing $\khat$ due to the neglect of global radial structure in
the analysis. Nevertheless, given the number of approximations made
in our analysis, the agreement is satisfactory, especially at large
$\khat$.

@@

Fig. \ref{lowfreq_eigenfunc} plots the eigenfunctions $W$ and $\delta 
v_z$ for the fundamental mode with $\khat=10$ (i.e. the mode with
maximum growth rate in Fig. \ref{iso_eigen_kx}). We find $W\propto z$
throughout most of the disk, corresponding to a constant vertical
velocity perturbation, as expected for the fundamental mode (see
Appendix \ref{iso_explicit}). We show a meridional visualization of
this mode in Fig. \ref{lowfreq_eigenfunc_2d}. The $x$ axis is
stretched for clarity; radial velocities are in fact typically much 
smaller than vertical velocities. Most of the meridional kinetic energy is
contained within $\sim 2H$ of the midplane due to the density 
stratification.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/eigenvectorW_iso} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  1cm]{figures/eigenvectorvz_iso}
  \caption{Pseudo-enthalpy perturbation $W$ (top) and vertical velocity
    perturbation $\delta v_z$ of the fundamental VSI with 
    $\khat=10$ in the fiducial disk with $\beta=10^{-3}$. The real 
    (imaginary) parts of $W$ and $\delta v_z$ are plotted as solid
    (dotted) lines. 
    \label{lowfreq_eigenfunc}
  }
\end{figure}

\begin{figure}
%  \includegraphics[scale=0.345,clip=true,trim=0cm 0cm 2.5cm
%  0cm]{figures/result2d_vel}\includegraphics[scale=0.345,clip=true,trim=1.9cm 0cm 0cm
%  0cm]{figures/result2d_mom} 
  \includegraphics[width=\linewidth]{figures/result2d_iso}
  \caption{Visualization of the fundamental VSI mode in
    Fig. \ref{lowfreq_eigenfunc}. The color scale corresponds to the
    pressure perturbation $\delta P=\rho W$ scaled by its maximum value.
    Arrows correspond to the vector field field $\sqrt{\rho}(\delta
    v_x,\delta v_z)$. Note that the horizontal axis has been stretched 
    for clarity.  
    \label{lowfreq_eigenfunc_2d}
  }
\end{figure}

\subsubsection{Case study: $\khat=10$}
In Fig. \ref{compare_modes_iso_kx10} we plot eigenvalues of unstable 
modes with $\khat=10$. The result for our fiducial setup is shown
in black dots. We also plot other calculations with different
vertical boundary conditions: rigid boundary and $\zmax=5H$ (red
crosses),  free surface and $\zmax = 3H$ (blue triangles), free surface
and $\zmax=7H$ (green crosses).

The fundamental mode, marked by `f', corresponds to the eigenvalue
with the smallest $|\sigma|$. We see that the fundamental mode is
robust against vertical boundary conditions as it appears at the same
position in all cases. Higher order modes with larger growth rates
(i.e. the branch where $\nu$ increases with $|\omega|$) require larger
vertical domains, since they have increasing number of nodes along 
$z$. The `tail' end with decreasing $\nu$ to the right
of the plot differ significantly between the cases, and are thus
likely associated with boundary conditions.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_modes_iso_kx10.ps}
  \caption{Unstable modes with $\khat=10$ in the fiducial disk model
    with $\beta=10^{-3}$. %  a disk with 
    % $(\gamma,\Gamma)=(1.4,1.011)$, $(p,q,\epsilon)=(-1.5,-1,0.1)$ and
    % $\beta=10^{-3}$. 
    The fiducial boundary condition is a free surface ($\Delta P=0$)
    with vertical domain size $\zmax=5H$  (black dots).  Other
    cases correspond to different boundary condtions applied at
    different heights. The fundamental mode is marked by `f'.   
    \label{compare_modes_iso_kx10} 
  }
\end{figure}

@@

To illustrate this, we plot the mode diagram for $\khat=30$ in  
Fig. \ref{compare_modes_iso_kx30} with different boundary 
conditions. Surface modes are readily identified for 
$\zmax\geq 5H$ and are marked by `s'. They typically
have the largest growth rates but require sufficiently large vertical
domain size and their eigenvalues vary with $\zmax$.

Due to their dependence on boundary conditions, we 
do not consider surface modes further in detail. These modes have
already been discussed in the aforementioned studies. In fact, we will
see in the next section that surface modes are more effectively
stabilized by finite thermal relaxation than the fundamental mode. %  Thus, surface 
% modes may not be the most relevant in realistic disks.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_modes_iso_kx30_tags.ps}
  \caption{Same as Fig. \ref{compare_modes_iso_kx10} but for $\khat=30$
    \label{compare_modes_iso_kx30}. Examples of surface modes are
    marked by `s'. 
  }
\end{figure}

\subsection{Effect of thermal relaxation}\label{therm_relax_eff}
We now consider the effect of finite thermal relaxation by increasing 
$\beta$. Note that for the standard disk model the critical thermal 
timescale $\beta_\mathrm{crit} = 0.125$.

Fig. \ref{compare_modes_cool_kx10} show unstable modes with $\khat=10$ 
and $\beta\in[0.01,0.1]$. In addition to our fiducial setup with 
$\zmax=5H$, we also plot results for $\zmax=3H$ and $\zmax=7H$. In all 
three cases the number of unstable modes decrease with increasing
$\beta$. The fundamental mode is not the most unstable for 
$\beta=0.01$ as there is a number of higher order modes, but 
their existence depends on $\zmax$.

However, the fundamental mode is dominant for $\beta \geq 
0.03$. Higher order modes, with larger growth rates, are more
effectively stabilized with increasing $\beta$. For $\beta=0.1$, which
is close to $\beta_\mathrm{crit}$, we only find the fundamental
mode. This is consistent with the discussion in 
\S\ref{iso_vsi_beta_crit}.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/compare_modes_cool_kx10_z3.ps} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/compare_modes_cool_kx10_z5.ps}
  \includegraphics[width=\linewidth]{figures/compare_modes_cool_kx10_z7.ps}
  \caption{Unstable modes with $\khat=10$ and thermal
    relaxation timescales $\beta=0.01$ (black dots), $\beta=0.03$ (red
    crosses), $\beta=0.05$ (green triangles) and $\beta=0.1$ (blue
    crosses); for different vertical domain sizes $\zmax=3H$ (top),
    $\zmax=5H$ (middle, our fiducial setup) and $\zmax=7H$
    (bottom). The fundamental mode is marked with `f'.  
    \label{compare_modes_cool_kx10} 
  }
\end{figure}

Fig. \ref{lowfreq_eigenfunc_cool}---\ref{lowfreq_eigenfunc_2d_cool}
shows the fundamental mode with $\khat=10$ in a disk with 
$\beta=0.1$. Comparing these plots with
Fig. \ref{lowfreq_eigenfunc}---\ref{lowfreq_eigenfunc_2d}, which have
$\beta=10^{-3}$, we see that increasing $\beta$ restricts the region 
characterized by $W\sim z$ and $\delta v_z\sim$ constant closer to the
mid-plane, with oscillatory behavior emerging near vertical boundaries
where buoyancy becomes important. This leads to the `tilted' pressure
perturbations seen in Fig. \ref{lowfreq_eigenfunc_2d_cool} 
(cf. Fig. \ref{lowfreq_eigenfunc_2d}). % However, most of the meridional
% momentum is still contained within $|z|\lesssim 2H$ as for 
% the case with $\beta=10^{-3}$.

@@

In Fig. \ref{compare_modes_cool_kx30} we plot eigenvalues for unstable
modes with $\khat=30$. In this case surface modes appear for
$\zmax\geq5H$. However, they are effectively stabilized with
increasing $\beta$. For $\zmax=3H$, the fundamental mode becomes
dominant for $\beta\geq 0.03$, as observed for $\khat=10$. For
$\zmax=5H$ and $\zmax=7H$, we find the emergence modes slightly more
unstable than the fundamental mode even when $\beta=0.1$ (the leftmost
eigenvalues in the lower two panels). However, these modes are clearly associated
with the vertical domain size as they are absent for $\zmax=3H$. 
% Thus, for $\zmax=5H$ and  
% $\zmax=7H$  the fundamental mode is likely dominant for 
% $\beta\geq0.05$

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/compare_modes_cool_kx30_z3.ps} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/compare_modes_cool_kx30_z5.ps}
  \includegraphics[width=\linewidth]{figures/compare_modes_cool_kx30_z7.ps}
  \caption{Same as Fig. \ref{compare_modes_cool_kx10} but for
    $\khat=30$. Examples of surface modes are marked with `s'. 
    \label{compare_modes_cool_kx30} 
  }
\end{figure}

Fig. \ref{compare_modes_cool_kx10} and \ref{compare_modes_cool_kx30} 
show that the fundamental mode is not sensitive to boundary
conditions. The fundamental mode also becomes dominant as $\beta$
approaches $\beta_\mathrm{crit}$.  
Thus, tracking the fundamental mode is a more robust way of 
assessing stability as $\beta$ is increased from zero, rather than
searching for the most unstable mode, which are affected by boundary 
conditions.

\subsection{Critical thermal relaxation
  timescale}\label{bcrit_num_test}
We now test the critical thermal relaxation timescale given by
Eq. \ref{iso_vsi_cond}. We begin from eigenvalues for the fundamental
mode with $\beta=10^{-3}$ and slowly increase the thermal
relaxation timescale up to $\beta=10^3$. Fig. \ref{bcrit_compare1}
show the obtained growth rates, together with the critical  
thermal relaxation timescale $\beta_\mathrm{crit}=0.125$ for our
fiducial disk parameters.

For $5\lesssim\khat\lesssim30$, we find $\beta_\mathrm{crit}$ provides
an accurate prediction of the upper limit to the thermal relaxation 
timescale for the fundamental mode. For smaller $\khat$, the VSI can
operate slightly beyond $\beta_\mathrm{crit}$ (recall that
$\beta_\mathrm{crit}$ was derived by assuming $\khat^2\gg 1$). For
$\khat\geq 50$, we find perturbations have large amplitudes near the
vertical boundaries, and complications to our numerical search of
eigenvalues arise from emergence of modes associated with boundary
conditions. Nevertheless, growth rates rapidly decrease as $\beta$
approaches and increase beyond $\beta_\mathrm{crit}$.

Fig. \ref{bcrit_compare1} is consistent with numerical simulations
performed by \cite{nelson13} for which a thermal relaxation timescale
$\beta\gtrsim 0.6$ stablizes the VSI. This can be anticipated from our
estimate of the critical thermal timescale $\beta_\mathrm{crit}\simeq
0.1$ for our fiducial disk parameters (which is the same as that in
\citeauthor{nelson13}). It is interesting to note that there is a
thermal relaxation timescale that maximizes the mode decay rate for
$1\lesssim \khat \lesssim 30$. Here, it is $\beta\simeq1$---$2$ or
about $0.1$---$0.3$ orbital periods. This is in fact consistent with 
Fig. 12 of \citeauthor{nelson13} where the perturbed kinetic energy in their
numerical simulations were observed to decay most rapidly for a
thermal timescale of $\sim 0.1$ orbits.

\begin{figure}
   \includegraphics[width=\linewidth]{figures/gcorr_compare2} 
   \caption{Growth rate of the fundamental VSI mode in the fiducial disk
     model as a function of the thermal relaxation timescale $\beta$
     The vertical line is the critical thermal timescale
     $\beta_\mathrm{crit}$ obtained 
     from Eq. \ref{iso_vsi_cond}. 
     \label{bcrit_compare1}}   
 \end{figure}

 We further test the dependence of the critical thermal relaxation timescale 
 $\beta_\mathrm{crit}$ on disk parameters
 (Eq. \ref{iso_vsi_cond}).  We use the previous case study as a
 reference point and vary $q\in[-0.2,-1.2]$, 
 $\gamma\in[1.2,2.0]$, and $\epsilon\in[0.02,0.1]$
 separately.
 % Here, we use a wavenumber $\khat=10$, for which we
 % find growth rates reach zero for the range of disk parameters
 % considered, so that a numerical $\beta_\mathrm{crit}$ can be defined precisely. 
 The numerically obtained $\beta_\mathrm{crit}$ is shown in
 Fig. \ref{bcrit_compare} in comparison with Eq. \ref{iso_vsi_cond}.
 Our numerical results generally agree with Eq. \ref{iso_vsi_cond}. The
 agreement improves with decreasing  $|q|$,  $\epsilon$ and increasing
 $\gamma$, i.e. weaker instability. There is a noticeable difference
 as $\gamma\to1$ because the derivation of $\beta_\mathrm{crit}$
 assumed $\gamma>1$.

 Fig. \ref{bcrit_compare}, together with Fig. \ref{bcrit_compare1} and 
 our finding that the fundamental mode becomes dominant as one
 increases $\beta$, shows that $\beta_\mathrm{crit}$ as given 
 by Eq. \ref{iso_vsi_cond} provides a  
 practical measure for the thermal relaxation timescale below which 
 VSI will operate.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.cm 0cm
  0cm]{figures/bcrit_compare_q.ps} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm
  0.8cm]{figures/bcrit_compare_g.ps}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm
  0.8cm]{figures/bcrit_compare_e.ps} 
  \caption{Dependence of the upper limit to the thermal relaxation timescale
    $\beta_\mathrm{crit}$ for the fundamental VSI mode on disk
    parameters. Our fiducial setup is $(\gamma, \Gamma= (1.4, 1.011)$
    and $(p,q,\epsilon)=(-1.5,-1,0.05)$. Top: varying 
    $q\in[-1.2,-0.2]$; middle: varying $\gamma\in[1.2,2.0]$; bottom:
    varying $\epsilon\in[0.02,0.1]$. The perturbation wavenumber is
    $\khat=10$.  
    \label{bcrit_compare}}  
\end{figure}

@@

For $\gamma>\Gamma$, i.e. stably stratified disks,
Fig. \ref{gcorr_compare_vnoniso} shows that introducing finite thermal
relaxation rapidly stabilizes the disk, similar to that observed for
nearly vertically isothermal disks (Fig. \ref{bcrit_compare1}). For
$\gamma=1.7,\,2.0,\,2.5$, growth rates reach zero at
$\beta\simeq0.17,\,0.083,\,0.045$, respectively. Interestingly, these
values are equal to $\epsilon|q|/(\gamma-\Gamma)$. This suggests that
the critical thermal relaxation timescale for vertically
non-isothermal disks can also be estimated by Eq. \ref{iso_vsi_cond}
but with $\gamma-1$ replaced by $\gamma-\Gamma$.
